Structuring College Access: The Market Segment Model and College Board Geomarkets

### 2025-07-08 text/manuscript-full-KS.qmd
revisions dallas rq1

The six Geomarkets within the Dallas-Fort Worth metropolitan area are illustrated in @fig-dallas-maps by total population across the three census periods. The Dallas-Fort Worth region represents a unique "metroplex" case formed by the convergence of two separate metropolitan areas, which creates a unique urban spatial structure that differs from the classic urban core and suburban ring model showcased by Chicago. As depicted in @fig-dallas-maps, two distinct inner-core Geomarkets (City of Dallas and City of Fort Worth) are seperated by a centrally located suburban Geomarket encompassing Irving, Arlington, & Grand Prairie. These central Geomarkets are surrounded by the more conventional outer ring of suburban Geomarkets (Dallas County, Collin and Rockwall Counties, and West of Dallas/Ft. Worth Metroplex).

From 1980 to 2020, the population of the overall Dallas-Fort Worth metropolitan area increased dramatically from approximately 2.7 Million to 6.8 Million. @fig-dallas-maps shows all six Geomarkets experienced population growth over the three census periods. However, the most pronounced increases occured in the West of Dallas/Fort Worth Metroplex and Collin & Rockwall Counties Geomarkets, which grew from approximately 340,000 to 1.4 Million and 170,000 to 1.2 Million, respectively. These population trends, as shown in @fig-dallas-maps, suggest population growth has been partly driven by urban sprawl from the metropolitan centers -- City of Forth Worth and City of Dallas -- into the surrounding suburban and rural areas of the metropolitan areas. This outward expansion may also account for the comparatively larger geographic size of the West of Dallas/Fort Worth Metroplex Geomarket from 1980 to 2000.

@@

@fig-dallas-rq1-race presents the racial and ethnic composition of each Geomarket within the Dallas-Fort Worth metropolitan area across the three census periods. In 1980, all Geomarkets were predominantly White, with White populations ranging from 62% in City of Dallas to nearly 92% in West of Dallas/Fort Worth Metroplex. By 2000, however, Geomarkets located in the central part of the metropolitan area exhibited a more balanced racial/ethnic distribution, including Dallas County, Irving, Arlington, and Grand Prairie, and City of Fort Wort. In contrast, Collin & Rockwall Counties and West of Dallas/Fort Worth Metroplex Geomarkets remained predominantly White in 2000 at 78% and 80%, respectively. By 2000, the City of Dallas was no longer a predominantly White Geomarket, with a population that was approximately 24% Black, 30% Hispanic, and 3% Asian. Over the next two decades, the Hispanic population grew steadily, contributing to greater diversity. By 2020, the City of Dallas Geomarket had the lowest proportions of White and Asian residents across the Dallas-Fort Worth Metropolitan area, with approximately 21% Black, 38% Hispanic, and 2% Asian residents.

These shifts in racial and ethnic composition also became more pronounced for other Geomarkets between 2000 and 2020. The Collin and Rockwall and West of Dallas/Fort Worth Metroplex Geomarkets experienced continued declines in their share of White residents. However, both Geomarkets remained predominantly White (58% and 63%, respectively). In contrast, all other Geomarkets experienced continued increases in the shares of Black and Hispanic residents, resulting in a diminished previously White-majority populations for Dallas County (33% White, 23% Black, 32% Hispanic; 9% Asian), Irving, Arlington, and Grand Prairie (30% White, 20% Black, 36% Hispanic; 12% Asian), and City of Fort Worth (46% White, 15% Black, 33% Hispanic; 4% Asian). <!--#  Lastly, continuous increases in the shares of Black and Hispanic residents further diversified the City of Dallas, making it the most racially/ethnically diverse Geomarket in the Dallas-Fort Worth metropolitan area. -->

@@

Socioeconomic demographics of median income, percent living below poverty, and educational attainment over time are presented in @fig-dallas-rq1-ses for the Dallas-Fort Worth metropolitan area. Overall, median income both within and across Geomarkets remained relatively stable across the three census periods. <!-- Between 1980 to 2000, affluence levels across Geomarkets were consistent, ranging from a low of \$82,000 in City of Fort Worth to a high of \$135,00 in the Collin & Rockwall Counties Geomarket. However, the substantial increase in median income for the Collin & Rockwall Counties Geomarket during this time period plateaued by 2020, reshuffling relatively affluence levels among Geomarkets. By 2020, Collin & Rockwall Counties remained the most affluent Geomarket with a median household income of \$133,000, followed by West of Dallas/Fort Worth Metroplex (\$116,000), and Dallas County (\$90,000). In contrast, median household income for City of Dallas and City of Fort Worth remained relatively stable through 2020. As a result, --> In 2020, both City of Dallas and City of Fort Worth continued to rank at the lowest levels of affluence within the metropolitan area at \$86,000 and \$87,000 \[MISMATCH BETWEEN GRAPH AND TABLE?\] median household incomes, respectively. By contrast, Collin & Rockwall Counties remained the most affluent Geomarket with a median household income of \$133,000, followed by West of Dallas/Fort Worth Metroplex (\$116,000). @fig-dallas-rq1-ses also illustrates changes in the percentage of residents living in poverty and educational attainment within Geomarkets over time, revealing similar dynamics across Geomarkets to those observed in median household income. Together, these trends underscore the growingly racial/ethnic diverse but widening economic inequality within the evolving Geomarket landscape of the Dallas-Fort Worth metroplex.

@@

```{r fig-dallas-rq1-race, echo = FALSE, fig.align = 'center', fig.cap = "Racial/Ethnic Composition of Dallas Area Geomarkets, 1980-2020", out.width = "100%"}
knitr::include_graphics('./../assets/images/rq1-dallas-race.png')
#knitr::include_graphics('./../../outputs/figures/enrollment_funnel.png')
#![The enrollment funnel](assets/images/enrollment_funnel.png)
```

```{r fig-dallas-rq1-ses, echo = FALSE, fig.align = 'center', fig.cap = "Socioeconomic Characteristics of Dallas Area Geomarkets, 1980-2020", out.width = "100%"}
knitr::include_graphics('./../assets/images/rq1-dallas-ses.png')
#knitr::include_graphics('./../../outputs/figures/enrollment_funnel.png')
#![The enrollment funnel](assets/images/enrollment_funnel.png)
```

>>>>

The six Geomarkets within the Dallas-Fort Worth metropolitan area are illustrated in @fig-dallas-maps by total population across the three census periods. The Dallas-Fort Worth region represents a unique "metroplex" case formed by the convergence of two separate metropolitan areas, which creates a unique urban spatial structure that differs from the classic urban core and suburban ring model showcased by Chicago. As depicted in @fig-dallas-maps, two distinct inner-core Geomarkets (City of Dallas and City of Fort Worth) are separated by a centrally located suburban Geomarket encompassing Irving, Arlington, & Grand Prairie. These central Geomarkets are surrounded by the more conventional outer ring of suburban Geomarkets (Dallas County, Collin and Rockwall Counties, and West of Dallas/Ft. Worth Metroplex).

From 1980 to 2020, the population of the overall Dallas-Fort Worth metropolitan area increased dramatically from approximately 2.7 Million to 6.8 Million. @fig-dallas-maps shows all six Geomarkets experienced population growth over the three census periods. However, the most pronounced increases occurred in the West of Dallas/Fort Worth Metroplex and Collin & Rockwall Counties Geomarkets, which grew from approximately 350,000 to 1.6 Million and 160,000 to 1.1 Million, respectively. These population trends, as shown in @fig-dallas-maps, suggest population growth has been partly driven by urban sprawl from the metropolitan centers -- City of Forth Worth and City of Dallas -- into the surrounding suburban and rural areas of the metropolitan areas. This outward expansion may also account for the comparatively larger geographic size of the West of Dallas/Fort Worth Metroplex Geomarket from 1980 to 2000.

@@

@fig-dallas-rq1-race presents the racial and ethnic composition of each Geomarket within the Dallas-Fort Worth metropolitan area across the three census periods. In 1980, all Geomarkets were predominantly White, with White populations ranging from 62% in City of Dallas to 93% in West of Dallas/Fort Worth Metroplex. By 2000, however, Geomarkets located in the central part of the metropolitan area exhibited a more balanced racial/ethnic distribution, including Dallas County, Irving, Arlington, and Grand Prairie, and City of Fort Wort. In contrast, Collin & Rockwall Counties and West of Dallas/Fort Worth Metroplex Geomarkets remained predominantly White in 2000 at 78% and 79%, respectively. By 2000, the City of Dallas was no longer a predominantly White Geomarket, with a population that was approximately 25% Black, 36% Hispanic, and 2% Asian. Over the next two decades, the Hispanic population grew steadily, contributing to greater diversity. By 2020, the City of Dallas Geomarket had the lowest proportions of White and Asian residents across the Dallas-Fort Worth Metropolitan area, with approximately 23% Black and 42% Hispanic residents.

These shifts in racial and ethnic composition also became more pronounced for other Geomarkets between 2000 and 2020. The Collin and Rockwall and West of Dallas/Fort Worth Metroplex Geomarkets experienced continued declines in their share of White residents. However, both Geomarkets remained predominantly White (57% and 63%, respectively). In contrast, all other Geomarkets experienced continued increases in the shares of Black and Hispanic residents, resulting in a diminished previously White-majority populations for Dallas County (31% White, 25% Black, 34% Hispanic; 8% Asian), Irving, Arlington, and Grand Prairie (29% White, 19% Black, 38% Hispanic; 11% Asian), and City of Fort Worth (42% White, 16% Black, 35% Hispanic; 4% Asian). <!--#  Lastly, continuous increases in the shares of Black and Hispanic residents further diversified the City of Dallas, making it the most racially/ethnically diverse Geomarket in the Dallas-Fort Worth metropolitan area. -->

@@

Socioeconomic demographics of median income, percent living below poverty, and educational attainment over time are presented in @fig-dallas-rq1-ses for the Dallas-Fort Worth metropolitan area. Overall, median income both within and across Geomarkets remained relatively stable across the three census periods. <!-- Between 1980 to 2000, affluence levels across Geomarkets were consistent, ranging from a low of \$82,000 in City of Fort Worth to a high of \$135,00 in the Collin & Rockwall Counties Geomarket. However, the substantial increase in median income for the Collin & Rockwall Counties Geomarket during this time period plateaued by 2020, reshuffling relatively affluence levels among Geomarkets. By 2020, Collin & Rockwall Counties remained the most affluent Geomarket with a median household income of \$133,000, followed by West of Dallas/Fort Worth Metroplex (\$116,000), and Dallas County (\$90,000). In contrast, median household income for City of Dallas and City of Fort Worth remained relatively stable through 2020. As a result, --> In 2020, both City of Dallas and City of Fort Worth continued to rank at the lowest levels of affluence within the metropolitan area at \$81,000 and \$86,000 median household incomes, respectively. By contrast, Collin & Rockwall Counties remained the most affluent Geomarket with a median household income of \$129,000, followed by West of Dallas/Fort Worth Metroplex (\$120,000). @fig-dallas-rq1-ses also illustrates changes in the percentage of residents living in poverty and educational attainment within Geomarkets over time, revealing similar dynamics across Geomarkets to those observed in median household income. Together, these trends underscore the growingly racial/ethnic diverse but widening economic inequality within the evolving Geomarket landscape of the Dallas-Fort Worth metroplex.

@@

```{r fig-dallas-rq1-race, echo = FALSE, fig.align = 'center', fig.cap = "Racial/Ethnic Composition of Dallas Area Geomarkets, 1980-2020", out.width = "100%"}
knitr::include_graphics('./../results/graphs/rq1/rq1_dallas_race.png')
#knitr::include_graphics('./../assets/images/rq1-dallas-race.png') [OLD RESULTS BASED ON ZIP GEOMARKETS]
#knitr::include_graphics('./../../outputs/figures/enrollment_funnel.png')
#![The enrollment funnel](assets/images/enrollment_funnel.png)
```

```{r fig-dallas-rq1-ses, echo = FALSE, fig.align = 'center', fig.cap = "Socioeconomic Characteristics of Dallas Area Geomarkets, 1980-2020", out.width = "100%"}
knitr::include_graphics('./../results/graphs/rq1/rq1_dallas_ses.png')

#knitr::include_graphics('./../assets/images/rq1-dallas-ses.png') [OLD RESULTS BASED ON ZIP GEOMARKETS]
#knitr::include_graphics('./../../outputs/figures/enrollment_funnel.png')
#![The enrollment funnel](assets/images/enrollment_funnel.png)
```
